Inferring generation-interval distributions from contact tracing data: \\ \emph{in prep, PRSB}

### 2020-05-24 doc/response.tex

No. The intrinsic distributions does not account for any spatial effect and therefore are different from the forward distributions.
We have tried to make this clear throughout the manuscript.
In particular, we provide the following sentence in Introduction to distinguish intrinsic from the realized distributions:

\newtext{Since not all infectious contacts result in infection, realized distributions differ from the intrinsic distribution.}

\revtext{
The "the backward generation interval (or the secondary generation time)" is statistically sound. An infectee, before being infected, has been exposed to many infectors in a prevalent cohort and
could be independently infected by any of them. Each infectee is only associated with a single backward generation time whose distribution arises
from an independent competing risk framework. At the population level,
all backward measured intervals are independent random variables (but
not necessarily identically distributed). On the other hand, the "forward
generation interval" is problematic. A typical infected individual may
infect several contacts, but not independently if the infectious period is
random. Therefore, it is not appropriate to consider that the times to
transmission of the pathogen to its infectees as an iid. sample. This can
be shown in Exercise 3.5 (f) of Yan and Chowell (2019). At the population level, should such data can be observed, there might exist correlated
intervals arising from the same infector, something to ponder about.
}

We appreciate this point and have tried not to imply strong assumptions, and also taken care to empirically validate our statistical approaches.

@@

\revtext{
Lines 59, 114-115, 129, etc.: There is a distinction between "truncation"
and "censoring" in survival analysis. In the current context (e.g. contact tracing), when the inclusion criteria is conditioning on the infection
event of the infectee and looking backward to ascertain the infection time
of its infector, it is right-truncation. It leads to over-sampling short durations (as stated in Line 129). I suggest to use "right-truncation" instead
of "censoring" throughout the paper, wherever it appears.
}

Done.

\revtext{
Lines 121 and 123: The backward generation-interval distribution $b_t(\tau)$ and
the "censored" (right-truncated?) generation-interval distribution $c_t(\tau)$
should be clearly defined and expressed (e.g. in terms of conditional distribution) before expressing the relationship (6).
}

We now define both backward and forward generation-interval distributions more appropriately in the text.
We now refer to $c$ as the ``aggregated'' generation-interval distribution (all realized generation intervals until certain time point).

@@

\revtext{
Lines 233-241: Once more, the sampling bias corresponds to right-truncation,
not censoring. A great deal of words is involved leading to gef $g_{eff}(\tau)$. The
mathematical language is more powerful and perhaps gef $g_{eff}(\tau)$ could be
explained more concisely (and take less space?)
}

Thank you so much for telling us to remove all examples of censoring and then continuing to tell us to remove specific examples of censoring! Also, censoring is very commonly used in this context, so you were being pedantic in the first place. But we removed it, and can you shut up now? \jd{Maybe try to tone this text down a little bit?}

We have shortened and tried to improve this explanation.

@@

\revtext{Lines 241-243: "In a large homogeneously mixing population, the effective
distribution is expected to be equivalent to the intrinsic distribution..."
Define both $g(\tau)$ and gef $g_{eff}(\tau)$ mathematically at the first place and prove
their equivalence in a large homogeneously mixing population.}

We now show that the initial forward generation-interval distribution is equivalent to the intrinsic distribution.
We also point to the analytical intractability of the spatial effects on the realized generation-interval distributions (and therefore the initial forward distribution):

@@

The likelihood function is provided in the Methods section.
\swp{TODO. It seems like they're different... Ugh. I'll get back to this.}

\revtext{The authors has used the notations $\RR$ (Line 35) and $\Ro$ (Lines 98-99) for reproduction numbers and
linked them via $\RR = \Ro S(t)$:}

\revtext{Shouldn't $\RR$ be a function of time $t$ via $S(t)$?}

We no longer use $\RR$ to avoid confusion. We now use $\Rini = \Ro S(0)$ to convey the idea that we consider the possibility that $S(0)$ may not be 1 but are not explicitly considering how $S$ changes through time.

@@

\revtext{Is $S(t)$ a number or a proportion? In deterministic models, this matters
less as one can argue that all deterministic models are describing the
change of proportions (by re-scaling on population size, assuming it is
constant); but it matters a lot in stochastic models because the population
size is an essential parameter.}

We have tried to clarify these points. $S$ is a proportion, but we do attempt to address issues of individuality appropriately.

\jd{Is the reviewer getting paid by the word, or what?}

@@

\revtext{
The authors stated that $\RR$ transcends to the final size of the epidemic
(with cited references). I boldly assume that the authors actually meant
by $\Ro$: This statement is in the context of a large outbreak that ends
when either there is no infectious individuals or there is no susceptible
individuals left in the population. Under a different context, the outbreak
may reach an endemic steady state with sustained on-going transmission
persisting for a very long period of time due to re-infection or demography.
It could be further said that $\Ro$ can transcend to the long term endemicity
level, such as $S(\infty) = \RR$: I leave to the authors to search the literature
for the latter
}

Yes, we meant $\Ro$. We have now clarified this in the text.
We now mention in Introduction that $\Ro$ can also affect the endemicity level and added a few citations.

\revtext{$g(\tau)$ as a p.d.f. given by (2) and Line 101 is also an equilibrium
distribution.}

While we agree with the reviewer that the intrinsic generation-interval can be interpreted an equilibrium distribution, this only holds under homogeneous mixing assumption. 
As we show in this paper, when there is spatial structure, the realized generation-interval distributions differ from the intrinsic one. 
Even when the entire population is susceptible, the forward generation-interval distribution of the single primary infector will still be subject to the egocentric effect.
Therefore, the realized generation intervals of a single, primary infector in an otherwise susceptible population will not follow the intrinsic generation-interval distribution. We do mention in the manuscript that the intrinsic distribution can be interpreted as an equilibrium distribution but do not want to over-emphasize this perspective, since we're interested in realized distributions on a spatial network.
The reviewer also presented a series of formulas on the mean generation interval for special cases of the SEIR model; these are interesting, but not a direction that we want to go with this version of the paper.

@@

\newtext{The \emph{intrinsic} generation-interval distributions describe the expected time distribution of infectious contacts made by a primary case.}

\revtext{The distribution $g(\tau)$ links the two intrinsic parameters R0 and r
(homogeneous mixing). Both $\Ro$ and $g(\tau)$ are intrinsic in the context of the equilibrium conditions. Therefore the parameter $r$ is a theoretical value and is very different from any observed exponential
or sub-exponential growth rate as the system is not at equilibrium.}

\revtext{Observed exponential or sub-exponential growth events are different from the infection events. The current paper only addresses the infection events (mostly unobservable). This was
already addressed in the comments with respect to Line 47-48.
Any attempt to link the estimated $r$ based on data collected
on observed events using the Euler-Lotka equation could be extremely misleading}

We have tried to clarify points related to this objection in the MS. 
Briefly, $r$ can be defined as the logarithmic rate of change in incidence $\dot i/i$. It changes through time, but we are interested in estimating the value near the beginning (after transients, but before appreciable depletion of susceptibles). To do this with (simulated or real) data, we use a method that estimates the initial value of $r$, while making a first-order correction for the fact that it might be changing.

@@

``Ma, Junling, Jonathan Dushoff, Benjamin M. Bolker, and David JD Earn. "Estimating initial epidemic growth rates." Bulletin of mathematical biology 76, no. 1 (2014): 245-260.''

We do expect the initial $r$, $R$ and forward generation time to obey Euler-Lotka.

@@

\revtext{
The intrinsic parameters R0 and r are parameters in branching
process, which is used to approximate the "initial phase" of the
outbreak and that approximation is poor. By the way, the condition in Lines 107-108 $S(t) = S$ is also inaccurate. The expected exponential growth is under the approximation $S(t)/n \approx 1$ and
$n \to \infty$; where $n$ is the population size. The so-called initial
phase (when the depletion of the susceptible population is negligible) totally ignores the difference between the initial disease free equilibrium and the highly explosive dynamic growth
}

The reviewer is apparently conflating the general case, which would in fact be problematic, with the case we consider: where the \emph{initial} population is not fully susceptible. In the latter case, there is a period where incidence is low and the depletion of the susceptible population is negligible.
To avoid confusion, we now use the term ``initial reproduction number'': $\Rini = \Ro S(0)$.

\revtext{
 It is too strong a statement that "the Erlang distributions are
more realistic than the exponentially distributed periods in the standard
SEIR model." The advantage of using the Erlang distributions is that the
system can be analyzed using ordinary differential equations. They are
only suitable for latent and infectious periods with very small variances
The Erlang distribution is not only a subset of the gamma distribution with integer valued shape parameters, but also it is limited to distributions with variances smaller than the corresponding exponential distribution with the same mean value.
}

The exponential distribution is a special case of the Erlang distribution (when the shape parameter is equal to 1). So our statement is at least technically correct.
We agree that we should not over-emphasize the flexibility of the Erlang distribution.
We now say ``often'' more realistic.

@@

We had omitted a sentence previous --- we have now clarified this sentence as follows:

\newtext{In the main text, we present results based on exponentially distributed latent and infectious periods;
we show results based on more Erlang distributed latent periods ($n_E=2$), which better matches the incubation period distribution of Ebola virus disease, in Appendix.}

\revtext{
It is more convenient to use the Laplace transforms as a tool.
}

The reviewers discussion of and suggestions about Laplace transformations are interesting, but we think beyond the scope of this paper. Our experience has been that Laplace transformations and moment generating functions are as likely to obscure the key concepts underlying disease dynamics as to clarify them.

\jd{I would skip the following text.}

### 2020-05-26 doc/network.tex
changes

\begin{flushleft}{
	\Large
	\textbf\newline{
		Inferring the generation-interval distribution that correctly links speed and strength of an epidemic
	}
}
\newline
\\
Sang Woo Park\textsuperscript{1,2,*}
David Champredon\textsuperscript{3,4}
Jonathan Dushoff\textsuperscript{3,5}
\\

\bigskip
\textbf{1} Department of Mathematics \& Statistics, McMaster University, Hamilton, Ontario, Canada
\\
\textbf{2} Department of Ecology and Evolutionary Biology, Princeton University, Princeton, New Jersey, USA
\\
\textbf{3} Department of Biology, McMaster University, Hamilton, Ontario, Canada
\\
\textbf{4} Agent-Based Modelling Laboratory, York University, Toronto, Ontario, Canada
\\
\textbf{5} Michael G. DeGroote Institute for Infectious Disease Research, McMaster University, Hamilton, Ontario, Canada
\\
\bigskip

*swp2@princeton.edu
\end{flushleft}

@@

Generation intervals, defined as the time between when an individual is infected and when that individual infects another person, link two key quantities that describe an epidemic: the reproductive number, $\RR$, and the rate of exponential growth, $r$.
Generation intervals are often measured through contact tracing by identifying who infected whom.
We study how observed intervals differ from ``intrinsic'' intervals that could be estimated by tracing individual-level infectiousness, and identify both spatial and temporal effects, including truncating (due to observation time), and the effects of susceptible depletion at various spatial scales. 
Early in an epidemic, we expect the variation in the observed generation intervals to be mainly driven by the truncating and the population structure near the source of disease spread; 
therefore, we predict that correcting observed intervals for the effect of temporal truncation but \emph{not} for spatial effects will provide a spatially informed ``effective'' generation-interval distribution, which will correctly link $r$ and $\RR$.
We develop and test statistical methods for temporal corrections of generation intervals,
and confirm our prediction using individual-based simulations on an empirical network.

@@

Realized generation intervals are also affected by spatial structure.
In a population that does not mix homogeneously, susceptibility will tend to decrease more quickly in the neighbourhood of infected individuals than in the general population.
This means that infectious contacts made late in an individual's infection are more likely to be ineffective due to contacts that were made earlier (because the contactee may have been infected already).
As a result, realized generation intervals (from the perspective of an infector) will typically have shorter mean than the intrinsic generation-interval distribution in a non-homogenous population.
This perspective allows us to reinterpret the finding of \cite{trapman2016inferring} that, given an intrinsic generation interval and an observed growth rate, the reproductive number on various network structures is always smaller on a network than would be predicted from homogeneous mixing.

While an epidemic is ongoing, realized generation intervals until the current epidemic time can be aggregated to form a single distribution.
While realized generation intervals can be difficult to measure in practice (because it is not possible to observe the exact moment at which individuals become infected), the aggregated distribution is analogous to data collected through contact tracing during an ongoing epidemic:
we typically want to try to make inference based on all available data.
These aggregated measurements are ``truncated'' because we do not know what happens after the current epidemic time.
These truncated intervals are similar to backward intervals during the exponential growth phase: there is a bias towards over-sampling shorter intervals, which are more likely to have concluded by the current epidemic time.
These observations allow us to make the following prediction: removing the truncation bias from aggregated generation intervals early in an epidemic will yield the initial forward generation-interval distribution, which is ``spatially corrected'' (i.e., contains information about the population structure) and will allow us to correctly infer the initial reproductive number from the exponential growth rate.

In this study, we explore spatiotemporal variation in realized (particularly forward) generation intervals.
We extend previous frameworks to study how the aggregated generation-interval distributions change over time and compare this distribution with the backward generation-interval distribution.
We classify spatial effects on realized generation intervals into three levels (egocentric, local, and global) and discuss how they affect realize generation-interval distributions.
Finally, we compare two methods for accounting for temporal bias and test our prediction using simulations.

@@

Assuming that the individual properties are independent of risk of infection, the population-level kernel is given by integrating over these individual variations:
\begin{equation}
K(\tau) = \int k (\tau; a) f(a) da,
\end{equation}
where $f(a)$ represents a probability density over a (possibly multi-dimensional) aspect space.
The population-level kernel describes the rate at which infectious contacts are made by an an infected individual, on \emph{average}.

Assuming that a population mixes homogeneously, we can write: 
\begin{equation}
K(\tau) = \RR_0 g(\tau),
\end{equation}
where $\RR_0 = \int K(\tau) d\tau$ is the basic reproductive number (the expected number of secondary cases caused by a randomly chosen infectious individual in a fully susceptible population \citep{anderson1991infectious}), 
and $g(\tau)$ is the expected time distribution of infectious contacts made by a primary case (the intrinsic generation-interval distribution \citep{champredon2015intrinsic}).
Since all infectious contacts lead to infection in a disease-free equilibrium of a homogeneously mixing population, $g(\tau)$ also describes the expected time distribution of secondary cases caused by an average primary case in a fully susceptible population and can be treated as an ``equilibrium'' distribution of the corresponding homogeneous mixing model.

In a homogeneously mixing population, current disease incidence at time $t$, $i(t)$, is the product of the current infectiousness of individuals infected in the past and the current proportion of the population susceptible, $S(t)$.
\begin{equation}
i(t) = S(t) \int K(s) i(t-s) ds = \RR_0 S(t) \int g(s) i(t-s) ds.
\end{equation}
This model, referred to as the renewal equation, can describe a wide range of epidemic models \citep{heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral, wallinga2007generation, roberts2007model, Champredon2018equivalence}.
Over a period of time where the susceptible proportion $S(t) = S(0)$ remains roughly constant, we would expect approximately exponential growth in incidence $i(t)$; assuming $i(t) = i(0) \exp(r t)$ yields the Euler-Lotka equation \citep{lotka1907relation}, which provides a direct link between the exponential growth rate $r$ and the initial reproductive number $\Rini = \Ro S(0)$:
\begin{equation}
\frac{1}{\Rini} = \int g(s) \exp(-r s) ds.
\label{eq:euler}
\end{equation}
Under the homogeneous mixing assumption, the intrinsic generation-interval distribution $g(\tau)$ provides the correct link between $r$ and $\Rini$.

@@

Realized generation intervals can be measured either forward (from the perspective of a cohort of infectors) or backward (from the perspective of a cohort of infectees) in time \citep{champredon2015intrinsic, britton2019estimation}.
The forward generation-interval distribution $f_t(\tau)$ describes the infection time of \emph{infectees} caused by infectors who were infected at time $t$.
Similarly, the backward generation-interval distribution $b_t(\tau)$ describes the infection time of \emph{infectors} of infectees who were infected at time $t$.
For a single infector-infectee pair, both backward and forward \emph{measurements} should give the identical generation interval.
Therefore, the total number of infection that occurs at time $t + \tau$ caused by individuals infected at time $t$ can be expressed in terms of both the forward and backward generation-interval distributions:
\begin{equation}
i(t) f_t(\tau) = i(t+\tau) b_{t+\tau}(\tau).
\label{eq:equivalence}
\end{equation}

In a homogeneously mixing population, the forward and backward generation-interval distributions can be calculated exactly.
The density of new infections occurring at time $t+\tau$ caused by infectors who were infected at time $t$ is given by
\begin{equation}
i_{t}(t+\tau) = \RR_0 i(t) g(\tau) S(t+\tau)
\end{equation}
As shown in \cite{champredon2015intrinsic}, the forward generation-interval distribution, $f_t(\tau)$, is proportional to $i_{t+\tau}(t)$:
\begin{equation}
f_t(\tau) = \frac{g(\tau) S(t+\tau)}{\int_0^\infty g(\tau) S(t+\tau) d\tau}.
\label{eq:forward}
\end{equation}
In this case, the initial foward generation-interval distribution during the exponential growth phase (when $S(t) \approx S(0)$) is equivalent to the intrinsic generation-interval distribution and, therefore, provides the correct link between $r$ and \Rini.
Likewise, the density of new infections occurring at time $t$ caused by infectors who were infected at time $t-\tau$ is given by
\begin{equation}
i_{t-\tau}(t) = \RR_0 i(t-\tau) g(\tau) S(t)
\end{equation}
The backward generation-interval distribution, $b_t(\tau)$, is proportional to $i_{t-\tau}(t)$:
\begin{equation}
b_t(\tau) = \frac{i(t-\tau) g(\tau)}{\int_0^\infty i(t-\tau) g(\tau) d\tau}.
\end{equation}

During an outbreak, generation intervals cannot be measured for infection events that have not happened yet.
This effect is called ``right-truncation''.
Therefore, even if we aggregate all realized generation intervals over the course of an epidemic, their mean will be shorter than the intrinsic generation intervals.
The aggregated generation-interval distribution, by definition, is a weighted average of backward generation-interval distributions (weighted by incidence) up until the epidemic time $t$:
\begin{equation}
a_t(\tau) \propto \int_{-\infty}^t b_s(\tau) i(s) ds.
\end{equation}
The aggregated generation-interval distribution can be also equivalently expressed in terms of the forward generation-interval distributions:
\begin{equation}
a_t(\tau) \propto \int_{-\infty}^{t-\tau} f_s(\tau) i(s) ds.
\end{equation}
\eref{equivalence} confirms that both expressions are identical.

For a single outbreak, the mean aggregated generation interval will always be shorter than the mean intrinsic generation interval (Figure~\ref{fig:censor}).
There are two reasons for this phenomenon.
First, longer generation intervals are more likely to be missed due to right truncation (and short generation intervals are more likely to be realized).
In particular, if we assume that the initial forward generation-interval distribution remains constant ($f_t(\tau) \approx f_0(\tau)$) when an epidemic is growing exponentially ($i(t) = i(0) \exp(rt)$), 
the initial aggregated (or backward) generation-interval distribution is just the initial forward generation-interval distribution discounted by the rate of exponential growth \citep{britton2019estimation}:
\begin{equation}
a_0(\tau) = b_0(\tau) \propto f_0(\tau) \exp(-r\tau).
\label{eq:exp}
\end{equation}
A deterministic simulation confirms that the aggregated generation-interval distribution has the same mean as the backward generation-interval distribution during this period (Figure~\ref{fig:censor}).
Second, the decreasing number of susceptibles over the course of an epidemic makes long infections less likely to occur \citep{champredon2015intrinsic}.
Overall, we expect naively using the aggregated generation-interval distribution, to underestimate the reproductive number.

\begin{figure}[!ht]
\includegraphics[width=\textwidth]{../fig/temporal_effect.pdf}
\caption{\textbf{Temporal variation in the mean backward and aggregated generation interval.}
A deterministic Susceptible-Exposed-Infectious-Recovered (SEIR) model is simulated using Ebola-like parameters \citep{who2014ebola}: mean latent period $1/\sigma = 11.4 \textrm{ days}$, mean infectious period $1/\gamma = 5 \textrm{ days}$, and the basic reproductive number $\RR_0 = 2$. 
The backward and the aggregated mean generation interval are calculated over the course of an epidemic.
The dotted horizontal line represents the mean intrinsic generation interval.
}
\label{fig:censor}
\end{figure}

@@

Infected individuals may contact the same susceptible individual multiple times, but only the first effective contact gives rise to infection in a given individual (after this, they are no longer susceptible).
Therefore, we expect realized generation intervals to be shorter than intrinsic generation intervals, on average.
To explore the effect of multiple contacts on realized generation intervals, we relax our assumption that the population is homogeneous.
Instead, we assume that a disease spreads on a network;
infected individuals contact their ``acquaintances'' at random, but ``acquaintanceships'' are predetermined by the network structure before the beginning of an epidemic \cite{trapman2016inferring}.

We first consider the infection process from an ``egocentric'' point of view, taking into account infectious contacts made by a single infector.
We define the egocentric kernel as the rate at which secondary infections are realized by a single primary case with aspect $a$ in the absence of other infectors:
\begin{equation}
\hat{k}(\tau; a) = k(\tau; a) \exp \left(- \delta(a) \int_0^\tau k(s; a) ds\right),
\end{equation}
where $k(\tau; a)$ is the individual-level intrinsic kernel and $e^{- \delta(a) \int_0^\tau k(s; a) ds}$ is the probability that a susceptible acquaintance has not yet been contacted by a particular infected individual.
The dilution term, $\delta(a)$, models how contacts are distributed among susceptible acquaintances.

Throughout this paper, we assume that there is a constant per-pair contact rate $\lambda$ \cite{trapman2016inferring}.
In this case, the infectiousness of an individual $\RR(a) = \int k(s; a) ds$ is the product of the number of acquaintances $N(a)$, which can vary among individuals, and the contact rate $\lambda$; the dilution term is equal to the reciprocal of the number of acquaintances: $\delta(a) = 1/N(a)$.
This assumption can be relaxed by allowing for asymmetry \citep{trapman2016inferring} or heterogeneity \citep{ball1997epidemics, ball2002general} in contact rates; 
for brevity, we do not pursue these directions here.

>>>>

\begin{flushleft}{
	\Large
	\textbf\newline{
		Inferring generation-interval distributions from contact-tracing data
	}
}
\newline
\\
Sang Woo Park\textsuperscript{1,2,*}
David Champredon\textsuperscript{3,4}
Jonathan Dushoff\textsuperscript{3,5,$\dagger$}
\\

\bigskip
\textbf{1} Department of Ecology and Evolutionary Biology, Princeton University, Princeton, New Jersey, USA
\\
\textbf{2} Department of Mathematics \& Statistics, McMaster University, Hamilton, Ontario, Canada
\\
\textbf{3} Department of Biology, McMaster University, Hamilton, Ontario, Canada
\\
\textbf{4} Agent-Based Modelling Laboratory, York University, Toronto, Ontario, Canada
\\
\textbf{5} Michael G. DeGroote Institute for Infectious Disease Research, McMaster University, Hamilton, Ontario, Canada
\\
\bigskip

*swp2@princeton.edu\\
$\dagger$dushoff@mcmaster.ca
\end{flushleft}

@@

Generation intervals, defined as the time between when an individual is infected and when that individual infects another person, link two key quantities that describe an epidemic: the initial reproductive number, $\Rini$, and the initial rate of exponential growth, $r$.
Generation intervals can be measured through contact tracing by identifying who infected whom.
We study how realized intervals differ from ``intrinsic'' intervals that could be estimated by measuring individual-level infectiousness, and identify both spatial and temporal effects, including truncating (due to observation time), and the effects of susceptible depletion at various spatial scales.
Early in an epidemic, we expect the variation in the realized generation intervals to be mainly driven by the truncating and the population structure near the source of disease spread; 
therefore, we predict that correcting realized intervals for the effect of temporal truncation but \emph{not} for spatial effects will provide the initial forward generation-interval distribution, which is spatially informed and correctly links $r$ and $\Rini$.
We develop and test statistical methods for temporal corrections of generation intervals,
and confirm our prediction using individual-based simulations on an empirical network.

@@

Realized generation intervals are also affected by spatial structure.
In a population that does not mix homogeneously, susceptibility will tend to decrease more quickly in the neighbourhood of infected individuals than in the general population.
This means that infectious contacts made late in an individual's infection are more likely to be ineffective due to contacts that were made earlier (because the contactee may have been infected already).
As a result, realized generation intervals (from the perspective of an infector) will typically have shorter mean than the intrinsic generation-interval distribution in a non-homogeneous population.
This perspective allows us to reinterpret the finding of \cite{trapman2016inferring} that, given an intrinsic generation interval and an observed growth rate, the reproductive number on various network structures is always smaller on a network than would be predicted from homogeneous mixing.

While an epidemic is ongoing, realized generation intervals can be measured by identifying who infected whom and can be aggregated to form a single distribution.
In practice, realized generation intervals is difficult to measure because it is difficult to observe the exact moment at which individuals become infected;
in most cases, observable events are clinical (e.g. onset of symptoms).
However, it is still possible to directly measure generation intervals through contact tracing in some cases (e.g., a rabid dog biting another dog \citep{hampson2009transmission}),.
Thus, the aggregated distribution is analogous to data collected through contact tracing during an ongoing epidemic ---
we typically want to try to make inference based on all available data that have been gathered since the beginning of an epidemic.

These aggregated measurements are ``truncated'' because we do not know what happens after the current epidemic time.
These truncated intervals are similar to backward intervals during the exponential growth phase: there is a bias towards over-sampling shorter intervals, which are more likely to have concluded by the current epidemic time.
These observations allow us to make the following prediction: removing the truncation bias from aggregated generation intervals early in an epidemic will yield the initial forward generation-interval distribution, which contains information about the population structure and allows us to correctly infer the initial reproductive number from the initial exponential growth rate.

In this study, we explore spatiotemporal variation in realized (particularly forward) generation intervals.
We extend previous frameworks to study how the aggregated generation-interval distributions change over time and compare this distribution with the backward generation-interval distribution.
We classify spatial effects on realized generation intervals into three levels (egocentric, local, and global) and discuss how they affect realize generation-interval distributions.
Finally, we compare two methods for accounting for temporal bias and test our prediction using individual-based simulations.

@@

Assuming that the individual properties are independent of risk of infection, the population-level kernel is given by integrating over these individual variations:
\begin{equation}
K(\tau) = \int k (\tau; a) f(a) \,\mathrm{d}a,
\end{equation}
where $f(a)$ represents a probability density over a (possibly multi-dimensional) aspect space.
The population-level kernel describes the rate at which infectious contacts are made by an an infected individual, on \emph{average}.

Assuming that a population mixes homogeneously, we can write: 
\begin{equation}
K(\tau) = \RR_0 g(\tau),
\end{equation}
where $\RR_0 = \int K(\tau)\,\mathrm{d}\tau$ is the basic reproductive number (the expected number of secondary cases caused by a randomly chosen infectious individual in a fully susceptible population \citep{anderson1991infectious}), 
and $g(\tau)$ is the expected time distribution of infectious contacts made by a primary case (the intrinsic generation-interval distribution \citep{champredon2015intrinsic}).
If the proportion of susceptibles contacted is not changing (for example in a homogeneously mixed at the endemic equilibrium or when the number of cases is vanishingly small), $g(\tau)$ also describes the realized generation intervals.

In a homogeneously mixing population, current disease incidence at time $t$, $i(t)$, is the product of the current infectiousness of individuals infected in the past and the current proportion of the population susceptible, $S(t)$.
\begin{equation}
i(t) = S(t) \int K(s) i(t-s)\,\mathrm{d}s = \RR_0 S(t) \int g(s) i(t-s)\,\mathrm{d}s.
\end{equation}
This model, referred to as the renewal equation, can describe a wide range of epidemic models \citep{heesterbeek1996concept, diekmann2000mathematical, roberts2004modelling, aldis2005integral, wallinga2007generation, roberts2007model, Champredon2018equivalence}.
Over a period of time where the susceptible proportion $S(t) = S(0)$ remains roughly constant, we would expect approximately exponential growth in incidence $i(t)$; assuming $i(t) = i(0) \exp(r t)$ yields the Euler-Lotka equation \citep{lotka1907relation}, which provides a direct link between the initial exponential growth rate $r$ and the initial reproductive number $\Rini = \Ro S(0)$:
\begin{equation}
\frac{1}{\Rini} = \int g(s) \exp(-r s)\,\mathrm{d}s.
\label{eq:euler}
\end{equation}
Under the homogeneous mixing assumption, the intrinsic generation-interval distribution $g(\tau)$ provides the correct link between $r$ and $\Rini$.

@@

Realized generation intervals can be measured either forward (from the perspective of a cohort of infectors) or backward (from the perspective of a cohort of infectees) in time \citep{champredon2015intrinsic, britton2019estimation}.
The forward generation-interval distribution $f_t(\tau)$ describes the infection time of \emph{infectees} caused by a cohort infectors who were infected at time $t$.
Similarly, the backward generation-interval distribution $b_t(\tau)$ describes the infection time of \emph{infectors} for a cohort of infectees who were infected at time $t$.
For a single infector-infectee pair, both backward and forward \emph{measurements} should give the identical generation interval.
Therefore, the total number of infection that occurs at time $t + \tau$ caused by individuals infected at time $t$ can be expressed in terms of both the forward and backward generation-interval distributions:
\begin{equation}
i(t) f_t(\tau) = i(t+\tau) b_{t+\tau}(\tau).
\label{eq:equivalence}
\end{equation}

In a homogeneously mixing population, the forward and backward generation-interval distributions can be calculated exactly.
The density of new infections occurring at time $t+\tau$ caused by infectors who were infected at time $t$ is given by
\begin{equation}
i_{t}(t+\tau) = \RR_0 i(t) g(\tau) S(t+\tau)
\end{equation}
As shown in \cite{champredon2015intrinsic}, the forward generation-interval distribution, $f_t(\tau)$, is proportional to $i_{t+\tau}(t)$:
\begin{equation}
f_t(\tau) = \frac{g(\tau) S(t+\tau)}{\int_0^\infty g(\tau) S(t+\tau) \,\mathrm{d}\tau}.
\label{eq:forward}
\end{equation}
In this case, the initial forward generation-interval distribution $f_0(\tau)$ during the exponential growth phase (when $S(t) \approx S(0)$) is equivalent to the intrinsic generation-interval distribution $g(\tau)$ and, therefore, provides the correct link between $r$ and \Rini.
Likewise, the density of new infections occurring at time $t$ caused by infectors who were infected at time $t-\tau$ is given by
\begin{equation}
i_{t-\tau}(t) = \RR_0 i(t-\tau) g(\tau) S(t)
\end{equation}
The backward generation-interval distribution, $b_t(\tau)$, is proportional to $i_{t-\tau}(t)$:
\begin{equation}
b_t(\tau) = \frac{i(t-\tau) g(\tau)}{\int_0^\infty i(t-\tau) g(\tau)\,\mathrm{d}\tau}.
\end{equation}

During an ongoing epidemic, generation intervals cannot be measured for infection events that have not happened yet.
This effect is called ``right-truncation''.
Therefore, even if we aggregate all realized generation intervals by identifying who infected whom through contact tracing (assuming that infection events are observable), their mean will be shorter than the intrinsic generation intervals.
The aggregated generation-interval distribution, by definition, is a weighted average of backward generation-interval distributions (weighted by incidence) up until the epidemic time $t$:
\begin{equation}
a_t(\tau) \propto \int_{-\infty}^t b_s(\tau) i(s) \,\mathrm{d}s.
\end{equation}
The aggregated generation-interval distribution can be also equivalently expressed in terms of the forward generation-interval distributions:
\begin{equation}
a_t(\tau) \propto \int_{-\infty}^{t-\tau} f_s(\tau) i(s) \,\mathrm{d}s.
\end{equation}
\eref{equivalence} confirms that both expressions are identical.

For a single outbreak, the mean aggregated generation interval will always be shorter than the mean intrinsic generation interval (Figure~\ref{fig:censor}).
There are two reasons for this phenomenon.
First, longer generation intervals are more likely to be missed due to right truncation (and short generation intervals are more likely to be realized).
In particular, if we assume that the initial forward generation-interval distribution remains constant ($f_t(\tau) \approx f_0(\tau)$) when an epidemic is growing exponentially ($i(t) = i(0) \exp(rt)$), 
the initial aggregated (or backward) generation-interval distribution is just the initial forward generation-interval distribution discounted by the rate of exponential growth \citep{britton2019estimation}:
\begin{equation}
a_0(\tau) = b_0(\tau) \propto f_0(\tau) \exp(-r\tau).
\label{eq:exp}
\end{equation}
A deterministic simulation confirms that the aggregated generation-interval distribution has the same mean as the backward generation-interval distribution during this period (Figure~\ref{fig:censor}).
Second, the decreasing number of susceptibles over the course of an epidemic makes long infectious contacts less likely to result in infection \citep{champredon2015intrinsic}.
Overall, we expect naively using the aggregated generation-interval distribution, to underestimate the initial reproductive number.

\begin{figure}[!ht]
\includegraphics[width=\textwidth]{../fig/temporal_effect.pdf}
\caption{\textbf{Temporal variation in the mean backward and aggregated generation interval.}
A deterministic Susceptible-Exposed-Infectious-Recovered (SEIR) model is simulated using Ebola-like parameters \citep{who2014ebola}: mean latent period $1/\sigma = 11.4 \textrm{ days}$, mean infectious period $1/\gamma = 5 \textrm{ days}$, and the basic reproductive number $\RR_0 = 2$. 
The mean backward and aggregated generation interval are calculated over the course of an epidemic.
The dotted horizontal line represents the mean intrinsic generation interval.
}
\label{fig:censor}
\end{figure}

@@

Infected individuals may contact the same susceptible individual multiple times, but only the first infectious contact gives rise to infection in a given individual (after this, they are no longer susceptible).
Therefore, we expect realized generation intervals from an individual to have a smaller mean than their mean intrinsic generation interval.
To explore the effect of multiple contacts on realized generation intervals, we relax our assumption that the population is homogeneous.
Instead, we assume that a disease spreads on a network;
infected individuals contact their ``acquaintances'' at random, but ``acquaintanceships'' are predetermined by the network structure before the beginning of an epidemic \cite{trapman2016inferring}.

We first consider the infection process from an ``egocentric'' point of view, taking into account infectious contacts made by a single infector.
We define the egocentric kernel as the rate at which secondary infections are realized by a single primary case with aspect $a$ in the absence of other infectors:
\begin{equation}
\hat{k}(\tau; a) = k(\tau; a) \exp \left(- \delta(a) \int_0^\tau k(s; a) \,\mathrm{d}s\right),
\end{equation}
where $k(\tau; a)$ is the individual-level intrinsic kernel and $e^{- \delta(a) \int_0^\tau k(s; a) ds}$ is the probability that a susceptible acquaintance has not yet been contacted by a particular infected individual.
The dilution term, $\delta(a)$, models how contacts are distributed among the acquaintances.

Throughout this paper, we assume that there is a constant per-pair contact rate $\lambda$ \cite{trapman2016inferring}.
In this case, the infectiousness of an individual $\RR(a) = \int k(s; a) \,\mathrm{d}s$ is the product of the number of acquaintances $N(a)$, which can vary among individuals, and the contact rate $\lambda$; the dilution term is equal to the reciprocal of the number of acquaintances: $\delta(a) = 1/N(a)$.
This assumption can be relaxed by allowing for asymmetry \citep{trapman2016inferring} or heterogeneity \citep{ball1997epidemics, ball2002general} in contact rates; 
for simplicity, we do not pursue these directions here.
